Sparse Function-space Representation \\ of Neural Networks

### 2023-05-15 paper/main.tex
Sec. 3

% Math Macros
\newcommand{\MB}{\mbf{B}}
\newcommand{\MC}{\mbf{C}}
\newcommand{\MZ}{\mbf{Z}}
\newcommand{\MV}{\mbf{V}}
\newcommand{\MX}{\mbf{X}}
\newcommand{\MA}{\mbf{A}}
\newcommand{\MK}{\mbf{K}}
\newcommand{\MI}{\mbf{I}}
\newcommand{\MH}{\mbf{H}}
\newcommand{\T}{\top}
\newcommand{\vzeros}{\mbf{0}}
\newcommand{\vtheta}[0]{\mathbold{\theta}}
\newcommand{\valpha}[0]{\mathbold{\alpha}}
\newcommand{\vkappa}[0]{\mathbold{\kappa}}
\newcommand{\vbeta}[0]{\mathbold{\beta}}
\newcommand{\vlambda}[0]{\mathbold{\lambda}}
\newcommand{\diag}{\text{{diag}}}

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, linearizing a trained neural network around optimal weights. The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation (GGN)~\cite{botev2017practical}. While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. This linear model, linear in the weights, refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains unclear without resorting to subset methods. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\textbf{Function-space methods for sequential learning} 
Continual learning (CL) hinges on how to deal with the non-stationary training distribution, wherein the training process may overwrite previously learned parameters, causing catastrophic forgetting \citep{mccloskey1989catastrophic}. CL approaches fall into inference-based, rehearsal-based, and model-based methods (see \cite{parisi2019continual, de2021continual} for an overview). The methods in the first category usually tackle this problem with weight-space regularization, such as EWC~\citep{kirkpatrick2017overcoming}, SI~\citep{zenke17a}), or VCL~\citep{nguyen-tuongModel2009} that induce retention of previously learned information in the weights alone. However, encoding functional knowledge in the weights does not guarantee good quality of the predictions. Function-space regularization techniques~\cite{li2018lwf, benjamin2018measuring, titsias2019functional, buzzega2020dark, pan2020continual, rudner2022continual} address this by relying on training data subsets for each task to compute the regularization term. These methods bridge the gap between objective and rehearsal-based methods since they use a subset of the training data for each task to compute the regularization term. Recently proposed methods, \eg, DER~\citep{buzzega2020dark}, FROMP~\citep{pan2020continual}, and S-FSVI~\citep{rudner2022continual} achieve state-of-the-art performance among the objective-based techniques in several CL benchmarks.

\textbf{Uncertainty quantification in RL}
A key challenge in RL is balancing the trade-off between exploration and exploitation \cite{sutton2018reinforcement}.
That is, should an agent select actions that it knows will lead to high reward (exploitation), or should it
select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction to balancing this trade-off is to model the uncertainty associated with a learned transition dynamics model and use it to guide exploration.
Prior work has taken an expectation over the dynamic model's posterior \cite{deisenrothPILCO2011,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampled from it (akin to Thompson sampling but referred to as posterior sampling RL)
\todo{cite thopmson sampling paper dearden et al 1999}
\cite{osbandMoreEfficientReinforcement2013},
and used it to implement optimism in the face of uncertainty using upper confidence bounds \cite{curiEfficient2020}.
\todo{cite UCRL paper}
No single strategy has emerged as a go to method and we highlight that these strategies are only as good as their uncertainty estimates.
% Central to these decision-making strategies are the methods for quantifying uncertainty.
Previous approaches have used GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}.
However, there has been little leveraging

@@

\section{\our: Sparse function-space represenation of neural networks}
%\paragraph{GP in the dual parameters}
The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process $f_{\textrm{post}}$ that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
%
\begin{equation}  \label{eq:gp_pred}
  \myexpect_{p(f_i \mid\vy)}[f_i]= \vk_{\vx i}^\top \valpha \quad \text{and} \quad
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] = k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
%
where the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters are the vector $\valpha \in \R^{m}$ and vector $\vbeta \in \R^{m}$, defined as:
%
\begin{equation}
  \label{eq:dual_param}
  \alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f_i}\log p(y_i \mid f_i)]
  \quad \text{and} \quad
  \beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f_i f_i}\log p(y_i \mid f_i)] .
\end{equation}
%
The relationships specified above are valid for generic likelihoods, and involve no approximations since the expectation is under the exact posterior. This higlights the fact that our approximate inference technique that will follow, while usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.

@@

As demonstrated in \cite{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper see the significance of the dual variables for approximate inference, but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \vbeta_{\vu} =  \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\vbeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1}   \valpha_{\vu} , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\MZ \in \R^{M \times D} $  instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplac}\todo{these variables need another subscript for differentiate that from the values computed at the data points} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(n^3)$ to $\mathcal{O}(m^3)$, with $m \ll n$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

In contrast to our approach, \citet{immer2021improving} use a subset of the data points to make a sparse GP model to reduce the computational complexity. As a result, this is a less principled approach for building a sparse model since it ignores the dataset's information. It is important to highlight that the sparsifying process is independent of the approximate inference technique because the dual variables of \cref{eq:dual_param} are computed using \cref{eq-empirical-risk} and a Laplace approximation.
More complicated inference techniques, such as variational inference, could be used, but given its simplicity, in our experiments we used the Laplace-GGN approximation with the trained neural network. Furthermore, our dual variable view of approximate inference means we do not need to retrain a separate sparse GP model. \todo{shouldn't this sentence be clear from the context? we derive the mean/kernel function that define the process, then we have a sparse GP already}
\section{Sparse Function-space Sequential Learning}

>>>>

% Math Macros
\newcommand{\MB}{\mbf{B}}
\newcommand{\MC}{\mbf{C}}
\newcommand{\MZ}{\mbf{Z}}
\newcommand{\MV}{\mbf{V}}
\newcommand{\MX}{\mbf{X}}
\newcommand{\MA}{\mbf{A}}
\newcommand{\MK}{\mbf{K}}
\newcommand{\MI}{\mbf{I}}
\newcommand{\MH}{\mbf{H}}
\newcommand{\T}{\top}
\newcommand{\vzeros}{\mbf{0}}
\newcommand{\vtheta}[0]{\mathbold{\theta}}
\newcommand{\valpha}[0]{\mathbold{\alpha}}
\newcommand{\vkappa}[0]{\mathbold{\kappa}}
\newcommand{\vbeta}[0]{\mathbold{\beta}}
\newcommand{\MBeta}[0]{\mathbold{B}}
\newcommand{\vlambda}[0]{\mathbold{\lambda}}
\newcommand{\diag}{\text{{diag}}}

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, linearizing a trained neural network around optimal weights. The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. This linear model, linear in the weights, refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains unclear without resorting to subset methods. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\textbf{Function-space methods for sequential learning} 
Continual learning (CL) hinges on how to deal with the non-stationary training distribution, wherein the training process may overwrite previously learned parameters, causing catastrophic forgetting \citep{mccloskey1989catastrophic}. CL approaches fall into inference-based, rehearsal-based, and model-based methods (see \cite{parisi2019continual, de2021continual} for an overview). The methods in the first category usually tackle this problem with weight-space regularization, such as EWC~\citep{kirkpatrick2017overcoming}, SI~\citep{zenke17a}), or VCL~\citep{nguyen-tuongModel2009} that induce retention of previously learned information in the weights alone. However, encoding functional knowledge in the weights does not guarantee good quality of the predictions. Function-space regularization techniques~\cite{li2018lwf, benjamin2018measuring, titsias2019functional, buzzega2020dark, pan2020continual, rudner2022continual} address this by relying on training data subsets for each task to compute the regularization term. These methods bridge the gap between objective and rehearsal-based methods since they use a subset of the training data for each task to compute the regularization term. Recently proposed methods, \eg, DER~\citep{buzzega2020dark}, FROMP~\citep{pan2020continual}, and S-FSVI~\citep{rudner2022continual} achieve state-of-the-art performance among the objective-based techniques in several CL benchmarks. \looseness-2

\textbf{Uncertainty quantification in RL}
A key challenge in RL is balancing the trade-off between exploration and exploitation \cite{sutton2018reinforcement}.
That is, should an agent select actions that it knows will lead to high reward (exploitation), or should it
select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction to balancing this trade-off is to model the uncertainty associated with a learned transition dynamics model and use it to guide exploration.
Prior work has taken an expectation over the dynamic model's posterior \cite{deisenrothPILCO2011,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampled from it (akin to Thompson sampling but referred to as posterior sampling RL)
\todo{cite thopmson sampling paper dearden et al 1999}
\cite{osbandMoreEfficientReinforcement2013},
and used it to implement optimism in the face of uncertainty using upper confidence bounds \cite{curiEfficient2020}.
\todo{cite UCRL paper}
No single strategy has emerged as a go to method and we highlight that these strategies are only as good as their uncertainty estimates.
Previous approaches have used GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}.
%However, there has been little leveraging

@@

\section{\our: Sparse function-space represenation of neural networks}
%\paragraph{GP in the dual parameters}
The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process $f_{\textrm{post}}$ that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
%
\begin{equation}  \label{eq:gp_pred}
  \myexpect_{p(f_i \mid\vy)}[f_i]= \vk_{\vx i}^\top \valpha \quad \text{and} \quad
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] = k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
%
where the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters are the vectors $\valpha, \vbeta \in \R^{N}$, defined as:\looseness-1
%
\begin{equation}
  \label{eq:dual_param}
  \alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f_i}\log p(y_i \mid f_i)]
  \quad \text{and} \quad
  \beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f_i f_i}\log p(y_i \mid f_i)] .
\end{equation}
%
The relationships specified above are valid for generic likelihoods, and involve no approximations since the expectation is under the exact posterior. This higlights the fact that our approximate inference technique that will follow, while usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.

@@

As demonstrated in \cite{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper see the significance of the dual variables for approximate inference, but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i=1}^N  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \MBeta_{\vu} =  \sum_{i=1}^N \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\MBeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1} \valpha_{\vu}
   \quad \text{and} \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = k_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \MBeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\{\vz_j\}_{i=1}^M$, $\vz \in \R^{D}$, instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplace} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$, with $M \ll N$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

Unlike our approach, \citet{immer2021improving} employ a subset of the data points to construct a sparse GP model, reducing computational complexity. However, this method could be considered less principled for building a sparse model, as it ignores contribution from data. It is important to note that the sparsifying process is independent of the approximate inference technique because the dual variables of \cref{eq:dual_param} are computed using \cref{eq-empirical-risk} and a Laplace approximation.
More complicated inference techniques, such as variational inference, could be used, but given its simplicity, in our experiments we used the Laplace-GGN approximation with the trained neural network. Furthermore, our dual variable view of approximate inference means we do not need to retrain a separate sparse GP model.%\todo{shouldn't this sentence be clear from the context? we derive the mean/kernel function that define the process, then we have a sparse GP already}

\section{Sparse Function-space Sequential Learning}
